---
layout: page
element: lecture
title: Demand
date: September 12, 2018
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
library(xkcd)
library(tidyr)
library(dplyr)
library(here)
library(pander)
```

## Feedback

<h3 style="color: green;">Good: </h3>
* Videos were interesting and helpful
* Sapling exercises in class were very helpful!

<h3 style="color: red;">Needs improvement: </h3>
* Please explain Nash Equilibrium and Tit-for-Tat


## What is demand?

<iframe src="https://giphy.com/embed/V9GZaqlnX3dsc" width="480" height="384" frameBorder="0" class="giphy-embed" allowFullScreen></iframe><p><a href="https://giphy.com/gifs/candy-silly-demand-V9GZaqlnX3dsc">via GIPHY</a></p>


## How do we estimate demand? { .smaller }

* Often estimated with surveys of __willingness to pay__ (WTP), with which we can build up a __demand schedule__.

* __Demand Schedule__ - _A table of quantity demanded at different price levels._
* __Willingness to Pay__ - _The maximum amount a person is willing to pay for a quantity of a good (e.g. eBay max)._

```{r ds, echo=FALSE}
d <- data.frame(
  Price = c(1, seq(10,50, 10))
) %>% mutate(Quantity = 10000 - 500*Price^.5)
```

`r pander(d)`

## A market model { .smaller }

A model of a market consists of:

* __Demand curve__ - _A graphical representation of a demand schedule. A demand curve depicts how much quantity of a good would be demanded at different prices._
* __Supply curve__ - _A graphical representation of a supply curve, showing how much quantity of a good is supplied at different prices._
* __Demand and supply curve shifts__ - _A shift in demand or supply due to a change in income (for demand) or technology (for supply), or other factors._
* __Market equilibrium__ - _The point at which demand and supply is equal (i.e. the curves intersect)._
* __Changes in market equilibrium__ - _How the market equilibrium changes in response to shifts in demand and supply._

## Demand curves

```{r dc}

ggplot(data = d) + 
  #xkcdline(data = d, aes(x = min(d$x), xend = max(d$x), y = min(d$y), yend = max(d$y)), color = "red", size = 2) + 
  geom_line(aes(x = Quantity, y = Price), color = "red", size = 1, alpha = .8) +
  xkcdaxis(c(6000,15000), c(0,50)) + 
  ggtitle("Demand Curve for WSU Football Tickets") + 
  xlab("Quantity of Tickets") + ylab("Price") +
  theme(text = element_text(size = 16, family = "Humor Sans")) 

```


## Changes in demand

* __Movement along the demand curve__ - _Due to a change in price, we move along the demand curve from the old price to the new price._
* __Shift in the demand curve__ - _If demand shifts, then the entire curve moves to the left or the right._

_Review Practice Problem 1._


## Shifts in Demand { .smaller }

Shifts in demand can occur because of changes in:

* __Population__ - More people means more demand for goods, so the quantity demanded increases for any price point. 
* __Income__ - If a population becomes more wealthy overall, the demand curve can also shift (though also likely pivots).
* __Tastes__ - Preferences can change in response to cultural shifts, major events, or fashion.
* __Expectations__ - Expectations about the future can affect demand (e.g. an incoming hurricane increases bottled water purchases).
* __Number of consumers__ - A change in the number of consumers, such as when the drinking age was raised to 21.


## Graphing a shift in Demand

```{r dc2}
d$Quantity2 = d$Quantity * 1.5
names(d) <- c("Price", "Current", "50% More Students")
dplt <- gather(d, key = Variable, value = Quantity, -Price)

ggplot(data = dplt) + 
  #xkcdline(data = d, aes(x = min(d$x), xend = max(d$x), y = min(d$y), yend = max(d$y)), color = "red", size = 2) + 
  geom_line(aes(x = Quantity, y = Price, group = Variable, color = Variable), size = 1, alpha = .8) +
  xkcdaxis(c(6000,15000), c(0,50)) + 
  ggtitle("Demand curve for WSU football tickets with 50% more students") + 
  xlab("Quantity of Tickets") + ylab("Price") +
  theme(text = element_text(size = 16, family = "Humor Sans"),
        plot.title = element_text(size = 12)) 
```


## Substitutes and complements { .smaller }

* __Substitutes__ - Two goods are substitutes if a decrease in the price of one good __decreases__ demand for the other good.

_Examples include food away from home (FAFH) and food at home (FAH)._

* __Complements__ - Two goods are complements if a decrease in the price of one good __increases__ demand for the other good.

_Examples include gas prices and milk._

_Review Practice Problem 3._


## Normal and inferior goods { .smaller }

* __Normal goods__ - A normal good is one in which an increase in income __increases__ demand for that good.

_Examples include pretty much all goods._

* __Inferior goods__ - Goods are inferior if an increase in income __decreases__ demand for that good.

_Examples?_


## Wrapping Up

* Complete the [knowledge check](https://goo.gl/forms/LQa51bqKHODk1Xlf2) for today's participation point.

* Please provide [feedback](https://goo.gl/forms/8SiOwdi19kN8cINJ3) on today's lecture.
