---
layout: page
element: lecture
title: 'Interpretating Price Elasticity of Demand and Consumer and Producer Surplus'
date: October 05, 2018
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
library(xkcd)
library(tidyr)
library(dplyr)
```

## Pregame

* Writing Assignment 2 (__Due 10/10__): Submit via blackboard
* Mean exam score: 63%. This is 10 percentage points higher than for other ECONS101 classes!

## Price elasticity of demand

* __Price Elasticity of Demand__ - _The change in quantity demanded that results from a change in price._

$$\varepsilon = \frac{\Delta Q / Q}{\Delta P / P}.$$


## Two extremes of price elasticity of demand

* Price elasticity of demand is zero: Change in quantity demanded for a (any) change in price is zero.

* Price elasticity of demand is infinite: Change in quantity demanded for a (any) change in price is infinite.

## Zero elasticity

```{r zero}
d <- data.frame(x = .5, y = seq(0,1, length.out = 100))
ggplot(d) + 
  #xkcdline(aes(x = .5, xend = .5, y = 0, yend = 1, color = "red", size = 2)) + 
  geom_segment(aes(x = .5, xend = .5, y = 0, yend = 1), color = "red", size = .5) +
  xkcdaxis(c(0,1), c(0,1)) + 
  ggtitle("Perfectly inelastic demand (elasticity = zero)") + 
  xlab("Quantity demanded") + ylab("Price") +
  theme(text = element_text(size = 14, family = "Humor Sans")) 
```


## Infinite elasticity

```{r infinite}
d <- data.frame(x = .5, y = seq(0,1, length.out = 100))
ggplot(d) + 
  #xkcdline(aes(x = .5, xend = .5, y = 0, yend = 1, color = "red", size = 2)) + 
  geom_segment(aes(x = 0, xend = 1, y = .5, yend = .5), color = "red", size = .5) +
  xkcdaxis(c(0,1), c(0,1)) + 
  ggtitle("Perfectly elastic demand (elasticity = infinity)") + 
  xlab("Quantity demanded") + ylab("Price") +
  theme(text = element_text(size = 14, family = "Humor Sans")) 
```



## Surplus

__Consumer Surplus__ - _The difference between what a buyer would have been willing to pay and the price they pay._

__Producer Surplus__ - _The difference between the price the good is sold for and the seller's cost to produce._

## Surplus { .flexbox .vcenter }

<iframe width="560" height="315" src="https://www.youtube.com/embed/KWbi-Lq1P28?start=40" frameborder="0" allow="autoplay; encrypted-media" allowfullscreen></iframe>

## Surplus

* __What is Edward's surplus?__

> * Willingness to pay (WTP) - Price = 4000 - 3000 = 1000.

* __What is Vivian's surplus?__

> * Price - Willingness to sell (or just cost) = 3000 - 2000 = 1000.

## Consumer and producer surplus

* See in class notes or module for more depth as well as discussion of Yakima Basin Irrigation Project.


## Wrapping Up

* Complete the [knowledge check](https://goo.gl/forms/bVuvvNN94gZktiz13) for today's participation point.

* Please provide [feedback](https://goo.gl/forms/8SiOwdi19kN8cINJ3) on today's lecture.
