---
title: 'Writing Assignment 4: Monopsony and the Minimum Wage'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
library(here)
library(pander)
library(xkcd)
library(tidyr)
library(dplyr)
```


**DUE**: Wednesday, November 28, 2018 by 5pm Pacific Time via NETUTOR

_Essays are submitted to a plagarism aglorithm to check for copying. Make sure you submit your own work. Plagarism is a failing offense._

A debate has raged in economics for several decades as to whether the minimum wage negatively affects employment or not. Recall that economic theory suggests that in a perfectly competitive market, raising the minimum wage would result in businesses reducing their employment and shifting to using more equipment. However, analysis of employment levels after minimum wages are implemented has been mixed, with some studies finding a negative impact and others finding no impact. One recent paper by [Naidu, Suresh and Posner](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=3129221) (2018) suggests that the labor market is not competitive, and as a result employers can supress wages because there are many fewer "buyers" of labor (i.e. the firms) than there are "sellers" of labor (i.e. the workers).

Suppose the labor market is described by **Figure 1**. Given our discussion of monopsony and competitive markets, write an essay describing how a minimum wage law affects employment under a perfectly competitive market and constrast that with how it would affect employment under a monopsony market in which buyers hold market power. The essay should have three sections:

1. Use the figure and the labelled points in your essay, discuss:

    a. the equilibrium employment and wage level in a competitive market.
    b. the effect of a minimum wage in a competitive market.
    c. the equilibrium employment and wage level in a monopsony market. Is this equilibrium lower or higher than under competition?
    d. the effect of a minimum wage on wages and labor in a monopsony market. Do wages and employment increase or decrease?

2. Read the section "I.A.3 The Social Cost of Monopsony" (starting on pg. 15 of Naidue, Suresh, and Posner) and discuss the negative effects of monopsony on the labor market. Include both redistributive and waste effects, as well as the four additional effects they discuss.

3. Read the first two paragraphs of "I.A.4 Recent Developments: Evidence of Labor Market Concentration" (starting on pg. 17 of NSP). Explain why Cark and Krueger's research suggested that workers were paid less than their marginal product, and what that meant for whether the minimum wage would lead to employees being fired.

```{r echo=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(xkcd)
library(here)
library(pander)

x <- 0:100
d <- tibble(
  "Labor" = x,
  "Labor Supply" = .5*(x^1.5),
  "Labor Demand" = 1000 - 100*(x^.5),
  "Marginal Cost" = .9*(x^1.5)
)

d2 <- gather(d, key = "Key", value = "Wage", -Labor)
ggplot(data = d2) +
  geom_segment(aes(x = 48, xend = 48, y = 0, yend = 300), color = "deepskyblue4", linetype = 2, alpha = .9, size = .1) +
  geom_segment(aes(x = 0, xend = 48, y = 300, yend = 300), color = "deepskyblue4", linetype = 2, alpha = .9, size = .1) +
  geom_segment(aes(x = 0, xend = 48, y = 170, yend = 170), color = "deepskyblue4", linetype = 2, alpha = .9, size = .1) +
  
  geom_segment(aes(x = 59.5, xend = 59.5, y = 0, yend = 230), color = "deeppink4", linetype = 3, alpha = .9, size = .1) +
  geom_segment(aes(x = 0, xend = 59.5, y = 230, yend = 230), color = "deeppink4", linetype = 3, alpha = .9, size = .1) +

  annotate("text", x = 48, y = 340, label = "A")  +
  annotate("text", x = 50, y = 150, label = "B")  +
  annotate("text", x = 59.5, y = 270, label = "C")  +
  geom_line(aes(x = Labor, y = Wage, group = Key, color = Key)) + 
  ggtitle("Figure 1: A Monopsony Labor Market") +
  xlab("Labor") + ylab("Wage") +
  #theme(text = element_text(size = 16, family = "Humor Sans")) +
  xkcdaxis(c(0,100),c(0,1000))
```

_Note: The wikipedia page on monopsony provides a good initial reference on monopsony. It may be helpful for understanding the first part of the assignment._

**References**

Naidu, Suresh and Posner, Eric A. and Weyl, E. Glen, Antitrust Remedies for Labor Market Power (February 23, 2018). Harvard Law Review, Forthcoming; University of Chicago Coase-Sandor Institute for Law & Economics Research Paper No. 850; U of Chicago, Public Law Working Paper No. 665. Available at SSRN: https://ssrn.com/abstract=3129221 or http://dx.doi.org/10.2139/ssrn.3129221

## Grading Rubric
Your essay must not exceed one typed page with one-inch margins and 12-point font single spaced.  Please include your name, instructor’s name (Potter), course and section number 1 (i.e. EconS 101.01), and writing assignment (assignment 3) on the top of your assignment in 12-point font. Your essay will be assessed as either “Meets Expectations” or “Needs Improvement” on both the writing portion and economic analysis. If your essay needs improvement, you will be given feedback from NetTutor to help you revise it.  You can then re-submit the essay to NetTutor for another try, but you only get one second chance. If you take materials or quotes from sources, you must provide citations/references in a standard citation/reference format. The references do not count toward the one page limit. Here are the criteria your essay will be judged against:

| Criterion | Needs Improvement | Meets Expectations |
| --------- | ----------------- | ------------------ |
| Writing   | The writing is difficult to follow and/or poorly organized.  Transition sentences are absent or ineffective.  Typos and/or grammatical errors distract the reader. Source material citations/references needed, but are missing or incorrect. | Ideas are well-organized.  Transition sentences effectively connect one idea to the next.  The essay is free of typos and grammatical errors. Sources properly cited and referenced, any citation style is acceptable. |
| Application of economic analysis | Does not correctly identify labels in Figure 1 and implications of minimum wage under competitive and monopsony markets. Does not include all effects of monopsony or does not explain them well. Does not correctly identify why Card and Krueger's work suggested that a monopsony market exists. | Correctly identifies the points A, B, and C in the figure. Discusses all 6 points in part 2, and provides correct identification of Card and Krueger's implications. |

